Distinct snow algae communities along an elevational gradient

### 2019-11-02 manuscript/manuscript.Rmd
init, copied over from Casey.Rmd. Removing figs from this file, put in figures.Rmd

---
title: "Distinct snow algae communities along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
date: "`r format(Sys.time())`"
output: 
  word_document:
    # toc: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: /home/cengstro/projects/sa_rbcl/Casey1/tmp/word-template.docx
  pdf_document: 
    number_sections: yes
    # toc: yes
    # toc_depth: 4
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
bibliography: /home/cengstro/ownCloud/mendeley_library/bibtex_by_grp/casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate. Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date. We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada. We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m). Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae. Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface. Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome.

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>
<br>
<br>

@@

### *Run title*: Elevational patterns in snow algae communities

### *To be submitted to*: 1) Nat Comm, 2) Frontiers in Micb 3) Env Micb 4) Microbial Ecology

@@

### Morphologically diverse snow algae blooms across a variety of habitats

Snow algae was widespread in our study area, beginning on May _ when we first observed snow algae blooming in the trees. Over the course of the season, we collected 309 samples from 33 dates and 13 moutains. We did not observe pink snow above treeline until June 6, despite several visits above treeline before that (site data table). We observed snow algae until the first snows of winter, although by later in the season most of the snow had melted out at lower elevations.

@@

### Diverse *rbc*L reveals species-level snow algae taxonomy

### Snow algae community composition varies with date and elevation

This is supported by our 18S and cell count data. Although we found Chlainomonas

### High elevation, high diversity

### Green snow algae turns orange at the surface

@@

<!--

When determining authorship the following criteria should be observed:

 - Substantial contributions to the conception or design of the work; or the
   acquisition, analysis, or interpretation of data for the work; AND
 - Drafting the work or revising it critically for important intellectual
   content; AND
 - Final approval of the version to be published ; AND
 - Agreement to be accountable for all aspects of the work in ensuring that
   questions related to the accuracy or integrity of any part of the work are
   appropriately investigated and resolved.

Contributors who meet fewer than all 4 of the above criteria for authorship
should not be listed as authors, but they should be acknowledged.
(http://www.icmje.org/roles_a.html)

-->

CE, LQ, and KY designed this study. Samples collected by CE with assistance from KY and LQ. KY and CE prepped samples for sequencing. All analyses were completed by CE and KY. All authors discussed the results and contributed to the final manuscript. CE wrote the manuscript with major input from LQ and KY.
<!-- The statement about the authors and contributors can be up to several sentences -->
<!-- long, describing the tasks of individual authors referred to by their initials -->
<!-- and should be included at the end of the manuscript before the References -->
<!-- section. -->

@@

# Figures {-}

<br>
<br>

![Figure 1. t-SNE clustering of snow algae rbcL amplicon sequence variants (ASVs). Colored points correspond to genus and species level assignment, grey points were not assigned at or below genus level. Labelled ellipses show OTUs. Perplexity = 30, theta = 0.5](/home/cengstro/projects/sa_rbcl/figs/asv_tsne_fig.png)

<!-- todo: italics, underscores, ChlorKri5, ChlorMura5, put ChloroKri4 on top of Chloro3, so its obvious that these aren't part of both OTUs.-->

![Figure 2. Sample ordination plots. A. NMDS comparing sample similarity using OTUs, shaded by elevation. Vectors indicate relative contribution of taxa driving differences between samples. Stress = 0.183. B. NMDS showing weighted UniFrac distances between snow algae samples. Stress = 0.11.](/home/cengstro/projects/sa_rbcl/figs/sample_ordination.png){width=13.01cm, height=18.56cm}

<!-- add dom morph groupings for UniFrac to connect with heatmap.  adjust size.  -->
<!-- change axis labs of b, MDS to NMDS-->

![Figure 3. Relative abundance heatmap of snow algae OTUs in each sample. Weight of color corresponds to the relative abundance of that OTU within a sample. Samples (rows) are grouped by the dominant cell morphology within that sample as determined by light microscopy. Within each morphology group samples are ordered by elevation, from low on the bottom to high on top. Sample IDs follow the convention: letters represent the mountain, and the number represents the order in which it was collected. Samples collected on the same day are labelled as .1, .2, and so on.](/home/cengstro/projects/sa_rbcl/figs/heatmap.png){width=12.88cm, height=17.17cm}

<!-- add elev range, date range. dom cell morph text to grey. lump chlainomonas categories (let it be a surprise)-->
<!-- add treeline -->
```{r, echo = F, include=F}
# otu_heat2 %>% 
#   group_by(bloom_type) %>% 
#   summarise(min(elev_m), max(elev_m))
# otu_heat2 %>% 
#   group_by(bloom_type) %>% 
#   summarise(min(date), max(date))
```

![Figure 4. Select microphotographs of snow algae. All photos are taken with light microscopy, except (f) which was taken with DIC. Scale bars represent 30 $\mu$m. (a) H5 (b) S10.2. (c) S4. (d) K2. (e) S10.1. (f) G16.1. (g) N12.1. (h) P20.1.](/home/cengstro/projects/sa_rbcl/figs/fig4.png){width=13.51cm, height=17.49cm}
<!-- order by elev, add sax1 -->

<!-- ![Figure 5. Relative abundance barplots comparing community composition in surface orange snow and green snow below in three samples.] -->
<!-- choose photo, microphotos, rethink colors (make consistant w heatmap) -->

# Supplementary Data {-}

![Supplementary Figure 1. Maximum likelihood tree (IQTree) of snow algae rbcL ASVs. Colors correspond to OTU. Edges with bootstrap values < 80 are displayed as dotted lines.](/home/cengstro/projects/sa_rbcl/figs/ggtree.png)

![Supplementary Figure 2. MDS plot of rbcL ASVs, colored by taxa. Ellipses contain all ASVs within an OTU. Stress =  ](/home/cengstro/projects/sa_rbcl/figs/asv_mds.png)

<!-- redo with NMDS, include stress vals -->
##### pagebreak

![Supplementary Figure 3. Relative abundance barplots comparing community composition in surface orange snow and subsurface green snow. * designates depth and surface samples collected on different days.](/home/cengstro/projects/sa_rbcl/figs/draft1/depth.png)

<!-- ![Supplementary Figure 4. Photos showing location of *Chloromonas krienitzii* dominated samples.]() -->

<!-- word to template: *[:alpha:] to \*, and *[:space:] to } -->

>>>>

---
title: "Distinct snow algae communities along an elevational gradient"
author:
  - name: Casey B. Engstrom
    institute: SFU
  - name: Kurt M. Yakimovich
    institute: SFU
  - name: Lynne M. Quarmby
    institute: SFU
    email: quarmby@sfu.ca
institute:
  - SFU: Dept. Molecular Biology and Biochemistry, Simon Fraser University, 8888 University Drive, Burnaby, BC, V5A 1S6, Canada
output: 
  bookdown::word_document2:
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=templates/scholarly-metadata.lua'
      - '--lua-filter=templates/author-info-blocks.lua'
    reference_docx: templates/word-template.docx
  pdf_document: 
    fig_caption: yes
    number_sections: yes
csl: templates/nature-communications.csl
bibliography: casey1.bib
abstract: |
  *Abstract*: Snow algae grow in melting summer snowfields throughout polar and alpine regions, forming extensive blooms that can alter snow surface albedo and increase melt rate. Snow algae community composition is reported to be similar between snow algae blooms in different continents; here, we report community composition can vary widely from patch to patch on the same mountain, and varies with elevation and date. We used dual-marker *rbc*L and 18S amplicon high-throughput sequencing (HTS) and light microscopy to assess community composition in 33 snow algae samples from the Coast Range of British Columbia, Canada. We found lower elevation snow algae blooms (800 - 1400 m) were dominated by *Chloromonas*, while *Sanguina* was restricted to higher elevations (1400 - 2200 m). Samples from higher elevation had higher phylogenetic diversity, and included several potentially novel clades of algae. Low elevation sites dominated by *Chloromonas krienitzii* remained green throughout the season in subsurface samples, but developed orange pigment at the surface. Taken together, our results suggest that microscopy and 18S amplicon HTS alone may not be sufficient to resolve differences between snow algae communities, and that snow algae may be specially adapted to distinct micro-habitats within the melting snow microbiome. 
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

### *Run title*: Snow algae communities along elevational...

@@

## Morphologically diverse snow algae blooms across a variety of habitats

Snow algae was widespread in our study area, beginning on May _ when we first observed snow algae blooming in the trees. Over the course of the season, we collected 309 samples from 33 dates and 13 moutains. We did not observe pink snow above treeline until June 6, despite several visits above treeline before that (S2). We observed snow algae until the first snows of winter, although by later in the season most of the snow had melted out at lower elevations.

@@

## Diverse *rbc*L reveals species-level snow algae taxonomy

Our *rbc*L amplicon library was dominated by sequences closely related to *Chloromonas*, *Chlainomonas*, and *Sanguina* (Fig. 2).

60% and 30% of *rbc*L sequences were assigned to genus and species level, while only % and % of our 18S amplicon library was assigned respectively. However, our 18S library generally supported our rbcL findings: samples dominated by *Sanguina* were consistant between markers, however samples containing *Chlainomonas* rbcL were assigned to *Chloromonas* 18S, which could be due to lack of differentiation between the the two genera using 18S (Supp. Fig. X).

*Chloromonas* was the most diverse

Algae OTU8 has a best BLAST score of 88%, with top hits in both Trebouxiophyceae and Chlorophyceae.

Currently there are just two species *Chlainomonas* in reference databases, but our results suggest there may be genetically distinct populations *Chlainomonas* at higher elevations (Fig. 1, Supp. Fig. 1, Supp. Fig 2).

There were fewer *rbc*L ASVs in *Sanguina* than in *Chloromonas* and *Chlainomonas*, and there were no visible sub-clusters within *Sanguina* in our t-SNE or phylogenetic tree (Fig. 2, S).

## Snow algae community composition varies with date and elevation

This is supported by our 18S and cell count data. Although we found Chlainomonas (Fig. 3))

Sites at higher elevation had higher *rbc*L diversity.

## Green snow algae turns orange at the surface

@@

CE, LQ, and KY designed this study. Samples collected by CE with assistance from KY and LQ. KY and CE prepped samples for sequencing. All analyses were completed by CE and KY. All authors discussed the results and contributed to the final manuscript. CE wrote the manuscript with major input from LQ and KY.

@@
